---
title: "Project1"
author: "ME"
date: "07/06/2022"
output: html_document
---

I begin by loading the project-related dataset and examining the first rows

```{r, setting the work directory and loading data, echo=TRUE}

wd <- "C:/Users/Amos.Turin/OneDrive - Essex County Council/Documents/Learning/R/Reproducible Project 1"
setwd(wd)

df <- read.csv("activity.csv")

head(df)
```

## What is mean total number of steps taken per day?

Calculating the total number of steps taken per day

```{r, echo=TRUE}

df$date <- as.Date(as.character(df$date), "%Y-%m-%d")
daily_steps <- aggregate(df["steps"], by=df["date"], sum)
head(daily_steps)

```

Plot the histogram of daily total steps

```{r, histogram, echo=TRUE}

par(mar=c(4,6,4,6))
hist(daily_steps$steps, breaks=20,
     xlab="Total number of steps",
     main="Histogram of total steps per day")

```
Finally, I calculate the mean and median values of total daily steps

```{r, echo=TRUE}

medSteps <- median(daily_steps$steps, na.rm=TRUE)
meanSteps <- mean(daily_steps$steps, na.rm=TRUE)

print(paste("Median: ", medSteps))
print(paste("Mean: ", round(meanSteps,2)))
```

## What is the average daily activity pattern?

I calculate the average number of steps for each 5-minute interval over all dates
in the dataset. Then I plot the timeseries of intervals.

```{r, aggregate and plot the timeseries, echo=TRUE}

df1 <- df[-which(is.na(df$steps)),]
interval_means <- aggregate(df1["steps"], by=df1["interval"], mean)

plot(interval_means$interval, interval_means$steps, 
     type='l', 
     xlab = "5-minute Intervals",
     ylab = "Average number of steps per interval",
     main = "Average number of steps per 5-minute interval")

```

From the plot it appears that the highest average number of steps in a 5-minute 
interval is in one of the intervals between 800-900 minutes.

Let's confirm what the maximum is and its relevant interval

```{r, echo=TRUE}

interval_means[which.max(interval_means$steps),]

```

The highest number of average steps is in the 835 interval at 206.2 steps.

## Imputing missing values

Calculating the total number of rows with NAs

```{r, echo=TRUE}
print(paste("Total number of rows with NAs:", sum(is.na(df))))
```
I can confirm this result by looking calling summary() on my dataframe.

This is indeed the number of rows with NAs, all of which are in the steps column.

```{r, echo=TRUE}
summary(df)
```

Imputing NAs with the mean of their relevant 5-minute interval

```{r, histogram of data after mean imputation, echo=TRUE}
options(warn=-1)
library(dplyr)

new_df <- df %>%
      group_by(interval) %>%
      mutate_at("steps", function(x) replace(x, is.na(x), mean(x, na.rm=TRUE)))
      
new_daily_steps <- aggregate(new_df["steps"], by=new_df["date"], sum)

hist(new_daily_steps$steps,
     breaks=20,
     xlab = "Total number of steps",
     main = "Histogram of total number of daily steps")

```

What has happened to the Mean and Median number of total daily steps as a 
result of the missing value imputation?

As I have imputed using the mean values, the overall mean has not changed, and
the median is now equal to the mean.

```{r, echo=TRUE}
newMedSteps <- median(new_daily_steps$steps, na.rm=TRUE)
newMeanSteps <- mean(new_daily_steps$steps, na.rm=TRUE)

print(paste("Median: ", round(newMedSteps,2)))
print(paste("Mean: ", round(newMeanSteps,2)))
```

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
options(warn=-1)
library(lubridate)
new_df <- new_df %>%
      mutate(weekday = wday(date, label=TRUE)) %>%
      mutate(weekday = ifelse(weekday == "Sun" | weekday == "Sat","Weekend","Weekday"))
      
new_df$weekday <- as.factor(new_df$weekday)
```

Plot the average steps per interval by Weekend vs Weekday

```{r, plot the histogram of weekend vs weekday, fig.width=10, echo=TRUE}

wkdy <- subset(new_df, weekday == "Weekday")
wknd <- subset(new_df, weekday == "Weekend")
weekend_interval_means <- aggregate(wknd["steps"], by=wknd["interval"], mean)
weekday_interval_means <- aggregate(wkdy["steps"], by=wkdy["interval"], mean)

par(mfrow = c(2,1))
plot(weekday_interval_means$interval, weekday_interval_means$steps, 
     type='l', 
     ylab = "",
     xlab = "",
     ylim = c(0,250),
     main = "Average number of steps per 5-minute interval by Weekday / Weekend")
grid(NULL, NULL)
plot(weekend_interval_means$interval, weekend_interval_means$steps, 
     type='l', 
     ylab = "",
     ylim = c(0,250),
     xlab = "5-minute Intervals")
grid(NULL, NULL)

```




